%Version 2.1 April 2023
% See section 11 of the User Manual for version history

%%\documentclass[pdflatex,sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style
\documentclass[referee,pdflatex,sn-basic]{sn-jnl}% referee := double spacing

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%
%%%%

\usepackage[super]{nth}% for superscripts


\theoremstyle{thmstyleone}%
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
\newtheorem{proposition}[theorem]{Proposition}% 

\theoremstyle{thmstyletwo}%
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%

\theoremstyle{thmstylethree}%
\newtheorem{definition}{Definition}%

\raggedbottom
%%\unnumbered% uncomment this for unnumbered level heads

\begin{document}

\title[Article Title]{Conservative ortholog families in 114 plants 
reveal relevant species and functional clusters
% Protein diversity in plant proteomes: 
% Elucidating evolutionary relationships through amino acid analysis
}

%%=============================================================%%
%% Prefix	-> \pfx{Dr}
%% GivenName	-> \fnm{Joergen W.}
%% Particle	-> \spfx{van der} -> surname prefix
%% FamilyName	-> \sur{Ploeg}
%% Suffix	-> \sfx{IV}
%% NatureName	-> \tanm{Poet Laureate} -> Title after name
%% Degrees	-> \dgr{MSc, PhD}
%% \author*[1,2]{\pfx{Dr} \fnm{Joergen W.} \spfx{van der} \sur{Ploeg} \sfx{IV} \tanm{Poet Laureate} 
%%                 \dgr{MSc, PhD}}\email{iauthor@gmail.com}
%%=============================================================%%

\author*[1,2]{\fnm{Nicolás} \sur{López-Rozo}}\email{nicolaslopez@javerianacali.edu.co}

\author[3]{\fnm{Fabian} \sur{Tobar-Tosse}}

\author[1,2]{\fnm{Jorge} \sur{Finke}}

\author[1,2]{\fnm{Camilo} \sur{Rocha}}\email{camilo.rocha@javerianacali.edu.co}


\affil*[1]{\orgdiv{Department of Electronics and Computer Science}, \orgname{Pontificia Universidad Javeriana Cali}, \orgaddress{\street{Calle 18 No. 118-250}, \city{Cali}, \postcode{111021}, \state{Valle del Cauca}, \country{Colombia}}}

\affil[2]{\orgdiv{OMICAS program}, \orgname{Pontificia Universidad Javeriana Cali}, \orgaddress{\street{Calle 18 No. 118-250}, \city{Cali}, \postcode{111021}, \state{Valle del Cauca}, \country{Colombia}}}

\affil[3]{\orgdiv{Department of Health Sciences}, \orgname{Pontificia Universidad Javeriana Cali}, \orgaddress{\street{Calle 18 No. 118-250}, \city{Cali}, \postcode{111021}, \state{Valle del Cauca}, \country{Colombia}}}

%%================================%%
%% Sample for structured abstract %%
%%================================%%

\abstract{
\textbf{Main Conclusion} 
This study unveils a hierarchical structure in 114 plant 
proteomes based on KOG category frequencies, resembling 
phylogenetic taxonomy for two plant clusters and showing
differentiated trends in annotation distributions for 3 
groups of plants. Additionally, noticeable correlations 
between Phytozome and Gene Ontology annotations support 
reliable insights into plant speciation and adaptation 
mechanisms. 


\textbf{Abstract}
% Proteins are essential for the functioning of cells in living 
% organisms, including plants. This study examines the protein 
% sequences of 114 plants, as well as 
% the related annotations using the concept of Eukaryotic 
% orthologous groups (KOGs). Based on the KOG category counts 
% of annotated proteins, a hierarchical classification that 
% groups similar plants is carried out, providing insights 
% into their shared evolutionary histories and adaptation 
% strategies. We also compare Gene Ontology and KOG categories 
% and find a strong correlation between the two rankings, 
% supporting the reliability of our results. These findings have 
% important implications for understanding plant speciation, 
% interactions in their environment, and responses to changing 
% conditions.

Functional annotations in plant genomes are essential properties 
for defining adaptive mechanisms under different stress 
conditions. Plants, when compared to animals, exhibit a higher 
grade of diversification and sophistication by its non-locomotive 
capacity. However, multi-omics approaches have no confidence in 
some functional enrichment due to its fuzzy phenotypic 
association. Therefore, new approaches based on biological 
insights should be applied. In this study, an exploration of 
114 plant proteomes elucidate singularities in the 
functional prioritization based on orthologous genes and their 
meaning in plant adaptative properties. Results show a 
differentiated trend in the density of functional annotations 
for physiologically distant groups (Eudicots, Monocots, Algae). 
Furthermore, frequency of some functional categories show 
relevance at the task of plant classification. Finally, the 
frequency patterns of the mentioned groups are proposed as an 
insight on the frequency of functional groups for other 
organisms.
}
\keywords{
Bioinformatics,
Functional diversity,
Hierarchical classification,
Comparative analysis,
Gene Ontology}

\maketitle

\subsection*{Abbreviations}

KOG: euKaryotic Orthologous Groups

COG: Clusters of Orthologous Genes



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}
\input{s1intro}


\section{Materials and Methods}
\label{sec:method}
\input{s2methods}


\section{Results}
\label{sec:results}
\input{s3results}


\section{Discussion and Conclusion}
\label{sec:conclusion}
\input{s4conclusion}






\backmatter

\vspace{1in}

\textbf{Author contribution statement} Conceptualization: N.L.-R. and 
F.T.-T.; Methodology: N.L.-R. and F.T.-T.; Software: N.L.-R.; 
Validation: N.L.-R.; Investigation: N.L.-R.; 
Data curation: N.L.-R.; 
Writing---original draft preparation: N.L.-R.; 
Writing---review and editing: F.T.-T., J.F. and C.R.; 
Project administration: J.F and C.R.; 
Funding acquisition: J.F. and C.R.;
All authors have read and agreed to the published version of the 
manuscript.

\textbf{Acknowledgments}
We would like to thank prof. Oliver Ebenhöh, Camila Riccio and 
Mauricio Ramírez-Castrillón for their helpful comments and 
insights.

\textbf{Funding}
This work was partially funded by the ``OMICAS program:
Optimización Multiescala In-silico de Cultivos Agrícolas 
Sostenibles (Infraestructura y validación en Arroz y Caña de 
Azúcar)'' Scientific Ecosystem belonging to the Colombia 
Científica Program, sponsored by The World Bank, The 
Ministry of Science, Technology and Innovation (MINCIENCIAS), 
ICETEX, the Colombian Ministry of Education and the Colombian 
Ministry of Commerce, Industry and Tourism, under GRANT ID: 
FP44842-217-2018, OMICAS Award ID: 792-61187.

\textbf{Data availability}
The datasets used and/or analyzed during 
the current study are available from the corresponding author on 
reasonable request.


\section*{Declarations}

\textbf{Conflicts of interest}
The authors declare no conflicts of interest relevant to the content 
of this manuscript.

\textbf{Ethics approval and consent to participate}
Not applicable.

\textbf{Consent for publication}
Not applicable.

\textbf{Open Access}
This article is licensed under a Creative Commons 
Attribution 4.0 International License, which permits use, sharing, 
adaptation, distribution and reproduction in any medium or format, as long
as you give appropriate credit to the original author(s) and the source,
provide a link to the Creative Commons license, and indicate if changes
were made. The images or other third party material in this article are
included in the article's Creative Commons license, unless indicated
otherwise in a credit line to the material. If material is not included 
in the article's Creative Commons license and your intended use is not
permitted by statutory regulation or exceeds the permitted use, you will
need to obtain permission directly from the copyright holder. To view a
copy of this license, visit \url{http://creativecommons.org/licenses/by/4.0/}

\begin{appendices}

\section{Section title of first appendix}\label{secA1}

An appendix contains supplementary information that is not an essential part of the text itself but which may be helpful in providing a more comprehensive understanding of the research problem or it is information that is too cumbersome to be included in the body of the paper.

%%=============================================%%
%% For submissions to Nature Portfolio Journals %%
%% please use the heading ``Extended Data''.   %%
%%=============================================%%

%%=============================================================%%
%% Sample for another appendix section			       %%
%%=============================================================%%

%% \section{Example of another appendix section}\label{secA2}%
%% Appendices may be used for helpful, supporting or essential material that would otherwise 
%% clutter, break up or be distracting to the text. Appendices can consist of sections, figures, 
%% tables and equations etc.

\end{appendices}

%%===========================================================================================%%
%% If you are submitting to one of the Nature Portfolio journals, using the eJP submission   %%
%% system, please include the references within the manuscript file itself. You may do this  %%
%% by copying the reference list from your .bbl file, paste it into the main manuscript .tex %%
%% file, and delete the associated \verb+\bibliography+ commands.                            %%
%%===========================================================================================%%

\bibliography{refs}
%% if required, the content of .bbl file can be included here once bbl is generated
%%\input sn-article.bbl


\end{document}
